`r params$project_title`

### 2022-02-10 reports/rmarkdown/01_data_pre_processing.Rmd

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)

setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown")
```

@@

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(sva)
library(batchelor)
remotes::install_github("czarnewski/niceRplots",upgrade = "always",force = T)
library(niceRplots)
library(tidyverse)

@@

```{r, echo = FALSE}
PATH <- "../../" #Path to the data

@@

```{r, Functions}
# create "other" taxonomy
other_taxa.fun <- function(df) {
  other_taxa <- df %>%
  unite("taxa_other", Kingdom:Species, sep = ";", remove = F, na.rm = T) %>%
  mutate(taxa_other = ifelse(is.na(.$Species), paste0(.$taxa_other, ";other"), .$taxa_other)) %>%
  #mutate(taxa_other = str_replace(.$taxa_other, ";NA.+|;NA", ";other")) %>%
  mutate(taxa_other = str_extract(.$taxa_other, "([^;]+);([^;]+)$")) %>% # get the two last taxa levels
  mutate(taxa_other = sub('^(.*);(other)', '\\2;\\1', .$taxa_other)) # place "other" first
return(other_taxa)
}

# create genus lacto taxonomy
genus_lacto.fun <- function(df){
  genus_lacto <- df %>%
    mutate(Genus_lacto = .$Genus, .after=Seq_ID) %>% 
    mutate(Genus_lacto = ifelse(grepl("crispatus",.$Species), paste("L.", "crispatus/acidophilus"), 
                                ifelse(grepl("reuteri",.$Species), paste("L.", "reuteri/oris/frumenti/antri"),
                                  ifelse(grepl("gasseri",.$Species), paste("L.","gasseri/johnsonii/taiwanensis"),
                                    ifelse(grepl("murinus",.$Species), paste("L.","murinus/animalis/apodemi/salivarius"),
                                      ifelse(grepl("plantarum",.$Species), 
                                           paste("L.", "plantarum/fabifermentans/composti/paraplantarum/graminis/fuchuensis"),
                                        ifelse(grepl("Lactobacillus",.$Genus), paste("L.", .$Species), 
                                              as.character(.$Genus_lacto))))))) )
  return(genus_lacto)
}

aggregate.fun <- function(df, level){
  level <- enquo(level)
  aggregated <- df %>%
    dplyr::rename(Taxonomy = !!level) %>%
    unite(., "Full_taxonomy", Kingdom:Species, sep = ";") %>%
    dplyr::select(-any_of(c("Sequence", "Seq_ID", "Genus_lacto", "taxa_other", "Full_taxonomy"))) %>%
    group_by(Taxonomy) %>%
    summarise(across(where(is.numeric), sum)) %>%
    ungroup()
return(aggregated)
}
```

@@

```{r, echo = FALSE}
# Load metadata
dataset_names <- c("ASV_tissue_V3_B2.csv",   # Tissue, Boston run 2 (95 samples)
                     "ASV_tissue_V3_B1.csv", # Tissue, Boston run 1 (1 sample)
                     "ASV_CVL_V2_B1.csv", # CVL V2, Boston run 1 (27 samples)
                     "ASV_CVL_V2_B2.csv", # CVL V2, Boston run 2 (49 samples)
                     "ASV_CVL_V2_S.csv",  # CVL V2, CTMR         (62 samples)
                     "ASV_CVL_V3_B1.csv") # CVL V3, Boston run 1 (111 samples)

datasets_ <- map(dataset_names, ~read.csv( paste0(PATH,"data/",.x))) %>% set_names(., dataset_names)
metadata <- read.csv(paste0(PATH,"data/Clinical_visit_2_3_updatefeb22.csv"))
rownames(metadata) <- as.character(metadata$ID)
```

# Taxonomy agglomeration

```{r, echo = FALSE}
temp <- map(datasets_, ~.x  %>%
            select(Sequence, Seq_ID, Kingdom:Species, everything()) %>%
            filter(!(is.na(.$Kingdom))) %>% # Removes any kingdom being NA
            other_taxa.fun(.) %>% # create "other" taxonomy 
            genus_lacto.fun(.)  # create genus lacto taxonomy
            ) %>%  
        {. ->> temp_ } %>% 
        map(., ~aggregate.fun(., "Genus_lacto")) # agglomerate on genus/lacto spp. lvl.

# Extract taxonomic information to file
seq_taxa <- bind_rows(temp_, .id = "dataset") %>% select(1:12) %>% unique() 
#write.csv(seq_taxa, paste0(PATH,"results/","SeqID_to_gen-lacto_tax-oth",".csv"),row.names = F )

# make long format raw non agglom ASV files
# remove_from_CVL_V2 <- intersect(colnames(temp_[[5]]), colnames(temp_[[3]]))[-c(1:11)]
# names <- c("ASV_tissue_V3_Boston run2","ASV_tissue_V3_Boston run1", "ASV_CVL_V2_Boston run1", 
#                      "ASV_CVL_V2_Boston run2", "ASV_CVL_V2_Stockholm", "ASV_CVL_V3_Boston run 1") 
# ASV_long <- temp_ %>%
#   modify_at(., c("ASV_CVL_V2_C.csv"), ~select(.x, -any_of(remove_from_CVL_V2))) %>% 
#   map(., ~pivot_longer(.x, matches("^P\\d"), names_to = "ID", values_to = "Reads")) %>%
#   set_names(names ) %>%
#   bind_rows(., .id = "dataset") %>%
#   separate(., dataset, into=c(NA, "Sample_type", "Visit","Place_sequenced"), sep = "_", remove = F) %>% 
#   mutate(dataset = str_extract(.$dataset, "CVL_V\\d|tissue_V3")) %>% 
#   group_split(dataset, .keep=F) %>% set_names(c("tissue", "CVL_V2", "CVL_V3"))
# 
# saveRDS(ASV_long, paste0(PATH,"results/","ASV_all_data_long_format.RDS"))

#arrange all datasets to have the same rownames with all bacteria
common_microbes_fun <- function(x) {
  t <- bind_rows(temp) %>%
    select(Taxonomy) %>%
    unique() %>%
    left_join(x, by="Taxonomy") %>%
    mutate(across(Taxonomy, ~replace_na(.x, "Not assigned"))) %>%
    mutate(across(-Taxonomy, ~replace_na(.x, 0))) %>%
    column_to_rownames(var = "Taxonomy")
  return(t)
}
datasets_ <- map(temp, ~common_microbes_fun(.x))
lapply(datasets_,dim)

@@

# Loading data and metadata Paulos code

```{r, eval=F, echo = FALSE}
PATH <- "../../" #Path to the data
# Load metadata
metadata <- read.csv(paste0(PATH,"data/Clinical_visit_2_3_updatefeb22.csv"))
rownames(metadata) <- as.character(metadata$ID)
# Load datasets
dataset_names <- c("ASV_tissue_B2.csv",
                     "ASV_tissue_B1.csv",
                     "ASV_CVL_V2_B1.csv",
                     "ASV_CVL_V2_B2.csv",
                     "ASV_CVL_V2_C.csv",
                     "ASV_CVL_V3.csv")
#import datasets and return matrix with taxa-level resolution
datasets <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"data/",x))
  temp <- as.matrix(x[,c((ncol(x)-6):ncol(x))])
  temp[is.na(temp)] <- "NA"
  sps <- apply( temp[,6:7] ,1,function(x) paste(x,collapse ="_"))
  not_complete <- grepl("_NA",sps)

  taxa <- apply( temp ,1,function(x) paste(x,collapse ="_"))
  taxa[not_complete] <- sub("_NA.*","",taxa[not_complete])
  taxa[not_complete] <- sub(".*_","other ",taxa[not_complete])
  taxa[!not_complete] <- sps[!not_complete]
  taxa <- sub("_"," ",taxa)
  use <- !is.na(taxa) & (taxa != "NA")
  temp <- as.matrix( x[,-c(1,(ncol(x)-7):ncol(x))] )
  temp <- rowsum(temp[use,],taxa[use])
  colnames(temp) <- colnames(x)[-c(1,(ncol(x)-7):ncol(x))]
  return(temp)
})
names(datasets) <- dataset_names
lapply(datasets,dim)
lapply(datasets,colnames)

# datasets <- datasets[3] #in case you wanna test in 1 dataset
# Transform phyloseq data into normalized matricies ( log2[CP1K] )
# datasets <- lapply(datasets,function(x){
#     x <- t(log2( t(x) / colSums(x) * 1000 + 1))
# })
# lapply(datasets,dim)

# Extract taxonomic information to file
taxa_info <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"data/",x))
  temp <- as.matrix(x[,c((ncol(x)-6):ncol(x))])
  temp[is.na(temp)] <- "NA"
  return(t(temp))
})
taxa_info <- t(as.data.frame(taxa_info))
rownames(taxa_info) <- sub("other.","other ",rownames(taxa_info))
rownames(taxa_info) <- sub("[.][0123456789]","other ",rownames(taxa_info))
sps <- apply( taxa_info[,6:7] ,1,function(x) paste(x,collapse ="_"))
not_complete <- grepl("_NA",sps)
taxa <- apply( taxa_info ,1,function(x) paste(x,collapse ="_"))
taxa[not_complete] <- sub("_NA.*","",taxa[not_complete])
taxa[not_complete] <- sub(".*_","other ",taxa[not_complete])
taxa[!not_complete] <- sps[!not_complete]
taxa <- sub("_"," ",taxa)
use <- !is.na(taxa) & (taxa != "NA")
taxa[taxa == "NA"] <- "other"
rownames(taxa_info) <- taxa
taxa_info <- taxa_info[! duplicated(rownames(taxa_info)),]
write.csv2(taxa_info, paste0(PATH,"results/taxonomy.csv"),row.names = T)

@@

```{r, echo = FALSE}
datasets <- datasets_
names(datasets)

#No batch correction needed in tissue_V3 dataset, only add it to the other samples
datasets[["ASV_tissue_V3"]] <- cbind( datasets[["ASV_tissue_B2.csv"]],datasets[["ASV_tissue_B1.csv"]] )

@@

```{r, echo = FALSE}
#mypar(3,3)
par(mar=c(3, 3, 0, 0))
PC_raw <- prcomp(round(t(batch1[nonzeros, paste0("boston_",common_samples)]),4) )

@@

if(file.exists(paste0(PATH,"results/encoder.h5")) & 
  file.exists(paste0(PATH,"results/decoder.h5")) & 
  file.exists(paste0(PATH,"results/aen.h5")) ){
  encoder <- load_model_hdf5(paste0(PATH,"results/encoder.h5"))
  decoder <- load_model_hdf5(paste0(PATH,"results/decoder.h5"))
  aen <- load_model_hdf5(paste0(PATH,"results/aen.h5"))

@@

  save_model_hdf5(encoder, paste0(PATH,"results/encoder.h5") )
  save_model_hdf5(decoder, paste0(PATH,"results/decoder.h5"))
  save_model_hdf5(aen, paste0(PATH,"results/aen.h5"))
}

@@

samples_not_batch1 <- sort(colnames(batch2_decoded_data) [! colnames(batch2_decoded_data) %in% common_samples2])
datasets[["ASV_CVL_V2"]] <- as.data.frame.array(cbind( batch1[nonzeros,] , batch2_decoded_data[,samples_not_batch1] ))
datasets <- datasets[-c(1,2,3)]
colnames(datasets[["ASV_CVL_V2"]]) <- sub(".*_","",colnames(datasets[["ASV_CVL_V2"]]))
lapply(datasets,dim)
lapply(datasets,colnames)

@@

common_microbes.fun <- function(x) {
  xy <- rownames_to_column(x, var = "Taxa")
  d <- data.frame(Taxa = common_microbes) %>%
    left_join(., xy, by="Taxa") %>%
    mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>%
    column_to_rownames(var = "Taxa")
  return(d)
}
datasets <- map(datasets, ~common_microbes.fun(.x))
lapply(datasets,dim)

@@

# raw files
n <- c("ASV_CVL_V3", "ASV_tissue_V3", "ASV_CVL_V2")
map2(datasets[1:3],n,  ~write.csv(.x, file=paste0(PATH,"results/",.y,"_raw_counts.csv"),row.names = T) )

# normalized files
datasets[["ASV_CVL_V3.csv"]] <- round( t(log2( t(datasets[["ASV_CVL_V3.csv"]]) / colSums(datasets[["ASV_CVL_V3.csv"]]) * 1000 + 1 )),4)
datasets[["ASV_tissue_V3"]] <- round(t( log2( t(datasets[["ASV_tissue_V3"]]) / colSums(datasets[["ASV_tissue_V3"]]) * 1000 + 1 )),4)
datasets[["ASV_CVL_V2_raw"]] <- round(t( log2( t(datasets[["ASV_CVL_V2_raw"]]) / colSums(datasets[["ASV_CVL_V2_raw"]]) * 1000 + 1 )),4)
datasets[["ASV_CVL_V2"]] <- round( datasets[["ASV_CVL_V2"]] ,4)

write.csv( datasets[["ASV_CVL_V2"]] , paste0(PATH,"results/ASV_CVL_V2_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_tissue_V3"]] , paste0(PATH,"results/ASV_tissue_V3_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_CVL_V3.csv"]] , paste0(PATH,"results/ASV_CVL_V3_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_CVL_V2_raw"]] , paste0(PATH,"results/ASV_CVL_V2_normalized_NOT_batch_corrected.csv"),row.names = T )

batches_CVL2 <- ifelse( colnames(datasets[["ASV_CVL_V2_raw"]]) %in% sub(".*_","",colnames(batch1)),"boston","CTMR" )
batches_CVL2 <- t(t(setNames(batches_CVL2 , colnames(datasets[["ASV_CVL_V2_raw"]]))))
write.csv( batches_CVL2 , paste0(PATH,"results/batches_CVL2.csv"),row.names = T )
```

```{r, echo = FALSE}
trx <- read.csv(paste0(PATH,"data/TRX_rawcounts.csv"),row.names = 1)
trx <- rowsum(trx[,-c(1,2)],trx$symbol)
colnames(trx) <- sub("ML","",colnames(trx))
trx  <- trx[ rownames(trx) != "" , ]
trx <- trx[ rownames(trx) != "NA" , ]
colnames(trx) <- as.character(metadata$ID)[ match(colnames(trx) , as.character(metadata$PatID) ) ]
trx <- trx[ rowSums(trx > 5) >= 3 , ]
trx <- t( log2( t( trx ) / colSums(trx) * 1e6 + 1 ) )
write.csv( trx , paste0(PATH,"results/Tissue_RNAseq_V3_normalized.csv"),row.names = T )
```

>>>>

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)
```

@@

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(sva)
library(batchelor)
remotes::install_github("czarnewski/niceRplots",upgrade = "always",force = T)
library(niceRplots)

@@

```{r, echo = FALSE}
PATH <- "~/Desktop/NBIS/SMS_Projects/broliden_5325/data/" #Path to the data
PATH <- "../../data/" #Path to the data

@@

@@

```{r, echo = FALSE}
# Load metadata
metadata <- read.csv(paste0(PATH,"/Clinical_visit_2_3_updatesept28.csv"))
rownames(metadata) <- as.character(metadata$ID)

# Load datasets
dataset_names <- c("ASV_tissue_B2.csv",
                     "ASV_tissue_B1.csv",
                     "ASV_CVL_V2_B1.csv",
                     "ASV_CVL_V2_B2.csv",
                     "ASV_CVL_V2_C.csv",
                     "ASV_CVL_V3.csv")

#import datasets and return matrix with taxa-level resolution
datasets <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"/",x))
  temp <- as.matrix(x[,c((ncol(x)-6):ncol(x))])
  temp[is.na(temp)] <- "NA"
  sps <- apply( temp[,6:7] ,1,function(x) paste(x,collapse ="_"))
  not_complete <- grepl("_NA",sps)

  taxa <- apply( temp ,1,function(x) paste(x,collapse ="_"))
  taxa[not_complete] <- sub("_NA.*","",taxa[not_complete])
  taxa[not_complete] <- sub(".*_","other ",taxa[not_complete])
  taxa[!not_complete] <- sps[!not_complete]
  taxa <- sub("_"," ",taxa)
  use <- !is.na(taxa) & (taxa != "NA")

  temp <- as.matrix( x[,-c(1,(ncol(x)-7):ncol(x))] )
  temp <- rowsum(temp[use,],taxa[use])
  colnames(temp) <- colnames(x)[-c(1,(ncol(x)-7):ncol(x))]
  return(temp)
})
names(datasets) <- dataset_names
lapply(datasets,dim)
lapply(datasets,colnames)

@@

# Extract taxonomic information to file
taxa_info <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"/",x))
  temp <- as.matrix(x[,c((ncol(x)-6):ncol(x))])
  temp[is.na(temp)] <- "NA"
  return(t(temp))
})
taxa_info <- t(as.data.frame(taxa_info))
rownames(taxa_info) <- sub("other.","other ",rownames(taxa_info))
rownames(taxa_info) <- sub("[.][0123456789]","other ",rownames(taxa_info))
sps <- apply( taxa_info[,6:7] ,1,function(x) paste(x,collapse ="_"))
not_complete <- grepl("_NA",sps)
taxa <- apply( taxa_info ,1,function(x) paste(x,collapse ="_"))
taxa[not_complete] <- sub("_NA.*","",taxa[not_complete])
taxa[not_complete] <- sub(".*_","other ",taxa[not_complete])
taxa[!not_complete] <- sps[!not_complete]
taxa <- sub("_"," ",taxa)
use <- !is.na(taxa) & (taxa != "NA")
taxa[taxa == "NA"] <- "other"
rownames(taxa_info) <- taxa
taxa_info <- taxa_info[! duplicated(rownames(taxa_info)),]
write.csv2(taxa_info, paste0(PATH,"../results/taxonomy.csv"),row.names = T)
```

@@

```{r, echo = FALSE}
#No batch correction needed in tissue_V3 dataset, only add it to the other samples
datasets[["ASV_tissue_V3"]] <- cbind( datasets[["ASV_tissue_B2.csv"]],datasets[["ASV_tissue_B1.csv"]] )

@@

```{r, echo = FALSE}
mypar(3,3)
PC_raw <- prcomp(round(t(batch1[nonzeros, paste0("boston_",common_samples)]),4) )

@@

if(file.exists(paste0(PATH,"../results/encoder.h5")) & 
  file.exists(paste0(PATH,"../results/decoder.h5")) & 
  file.exists(paste0(PATH,"../results/aen.h5")) ){
  encoder <- load_model_hdf5(paste0(PATH,"../results/encoder.h5"))
  decoder <- load_model_hdf5(paste0(PATH,"../results/decoder.h5"))
  aen <- load_model_hdf5(paste0(PATH,"../results/aen.h5"))

@@

  save_model_hdf5(encoder, paste0(PATH,"../results/encoder.h5") )
  save_model_hdf5(decoder, paste0(PATH,"../results/decoder.h5"))
  save_model_hdf5(aen, paste0(PATH,"../results/aen.h5"))
}

@@

samples_not_batch1 <- sort(colnames(batch2_decoded_data) [! colnames(batch2_decoded_data) %in% common_samples2])
datasets[["ASV_CVL_V2"]] <- cbind( batch1[nonzeros,] , batch2_decoded_data[,samples_not_batch1] )
datasets <- datasets[-c(1,2,3)]
colnames(datasets[["ASV_CVL_V2"]]) <- sub(".*_","",colnames(datasets[["ASV_CVL_V2"]]))
lapply(datasets,dim)
lapply(datasets,colnames)

@@

datasets <- lapply(datasets, function(x) {
  temp <- matrix(0,nrow = length(common_microbes),
                 ncol = ncol(x),
                 dimnames = list(common_microbes,colnames(x)))
  temp[rownames(x),colnames(x)] <- x
  return(temp)
})
lapply(datasets,dim)

@@

datasets[["ASV_CVL_V3.csv"]] <- round( t(log2( t(datasets[["ASV_CVL_V3.csv"]]) / colSums(datasets[["ASV_CVL_V3.csv"]]) * 1000 + 1 )),4)
datasets[["ASV_tissue_V3"]] <- round(t( log2( t(datasets[["ASV_tissue_V3"]]) / colSums(datasets[["ASV_tissue_V3"]]) * 1000 + 1 )),4)
datasets[["ASV_CVL_V2_raw"]] <- round(t( log2( t(datasets[["ASV_CVL_V2_raw"]]) / colSums(datasets[["ASV_CVL_V2_raw"]]) * 1000 + 1 )),4)
datasets[["ASV_CVL_V2"]] <- round( datasets[["ASV_CVL_V2"]] ,4)

write.csv( datasets[["ASV_CVL_V2"]] , paste0(PATH,"../results/ASV_CVL_V2_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_tissue_V3"]] , paste0(PATH,"../results/ASV_tissue_V3_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_CVL_V3.csv"]] , paste0(PATH,"../results/ASV_CVL_V3_normalized_batch_corrected.csv"),row.names = T )
write.csv( datasets[["ASV_CVL_V2_raw"]] , paste0(PATH,"../results/ASV_CVL_V2_normalized_NOT_batch_corrected.csv"),row.names = T )

batches_CVL2 <- ifelse( colnames(datasets[["ASV_CVL_V2_raw"]]) %in% sub(".*_","",colnames(batch1)),"boston","CTMR" )
batches_CVL2 <- t(t(setNames(batches_CVL2 , colnames(datasets[["ASV_CVL_V2_raw"]]))))
write.csv( batches_CVL2 , paste0(PATH,"../results/batches_CVL2.csv"),row.names = T )
```

```{r, echo = FALSE}
trx <- read.csv(paste0(PATH,"/TRX_rawcounts.csv"),row.names = 1)
trx <- rowsum(trx[,-c(1,2)],trx$symbol)
colnames(trx) <- sub("ML","",colnames(trx))
trx  <- trx[ rownames(trx) != "" , ]
trx <- trx[ rownames(trx) != "NA" , ]
colnames(trx) <- as.character(metadata$ID)[ match(colnames(trx) , as.character(metadata$PatID) ) ]
trx <- trx[ rowSums(trx > 5) >= 3 , ]
trx <- t( log2( t( trx ) / colSums(trx) * 1e6 + 1 ) )
write.csv( trx , paste0(PATH,"../results/Tissue_RNAseq_V3_normalized.csv"),row.names = T )
```
